---
title: "preprocess"
author: "Jianing Yao"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
```

### Step 1: read data and clean
```{r}
# read data
# the spreadsheet has 19 metabolic measures and 5 retinal thickness
data <- as.data.frame(read_excel("./Data/Sachdeva_Final_datasheet_OCT_metabolic_232023.xlsx"))
```

```{r}
# convert id and gender to factors
df <- data
df$IRB <- as.factor(df$IRB)
df$gender <- as.factor(df$gender)
levels(df$gender) <- c("female", "male")
```

### Step 2: descriptive statistics
```{r}
# gender 
table(df$gender)
table(df$gender)/nrow(df)
```


## table 1: mean and sd of metabolic measures and retinal measures
```{r}
cols_ind <- c(4:112)
median_data <- apply(df[, cols_ind], 2, function(x) median(x, na.rm = TRUE))
mean_data <- apply(df[, cols_ind], 2, function(x) mean(x, na.rm = TRUE))
sd_data <- apply(df[, cols_ind], 2, function(x) sd(x, na.rm = TRUE))
summary <- cbind(median_data, mean_data, sd_data)
colnames(summary) <- c("Median", "Mean", "SD")
round(summary,2)
```


## histograms of independent variables (age + metabolic measures)

```{r}
colnames(df)
indep_list <- c("age", "heart_rate","MAP","SBP","BMI","HbA1c","BUN","Cr","BUN/Cr","HDL","LDL","Triglycerides","Cholesterol","NEFA","Glucose","Ketone","Insulin","Adiponectin","Leptin","RAGE")
for (col in indep_list) {
  hist(df[[col]], xlab = paste(col), main = paste("Histogram of", col))
}

```


## replace missing values in metabolic measures with median
```{r}
met_list <- c("heart_rate","MAP","SBP","BMI","HbA1c","BUN","Cr","BUN/Cr","HDL","LDL","Triglycerides","Cholesterol","NEFA","Glucose","Ketone","Insulin","Adiponectin","Leptin","RAGE")
df.imp <- df
for (col in met_list) {
  df.imp[is.na(df.imp[[col]]), col] <- median(df.imp[[col]], na.rm = TRUE)
}
```




## Pairs plot, correlations, heatmap of metabolic measures

```{r}
pairs(df.imp[met_list])
```

```{r}
met_cor <- cor(df.imp[met_list], use = "pairwise.complete.obs")
corrplot(met_cor, type = "upper", method = "circle")
```


## Correlation heatmaps for retinal measures by eye
```{r}
# right eye
OD_list <- c(23:67)
OD_cor <- cor(df.imp[OD_list], use = "pairwise.complete.obs")
corrplot(OD_cor, type = "upper", method = "circle", tl.cex = 0.5)
```

```{r}
# left eye
OS_list <- c(68:112)
OS_cor <- cor(df.imp[OS_list], use = "pairwise.complete.obs")
corrplot(OS_cor, type = "upper", method = "circle", tl.cex = 0.5)
```


```{r}
# save cleaned dataset
write.table(df.imp, "./Data/data_cleaned.txt", sep = "\t", row.names = FALSE, col.names = TRUE)
```




TODO:
1. histograms for retinal thickness 
2. descriptive by 3 groups?
3. missing values for metabolic measures (REPLACE WITH MEDIAN FOR NOW)
4. A heatmap showing correlations between selected metabolic parameters and retinal thickness values in each of the anatomical locations to select predictors.  




Questions: 
1. gender code: which is female vs male



